\RequirePackage{silence} % :-\
  \WarningFilter{scrreprt}{Usage of package `titlesec'}
  \WarningFilter{titlesec}{Non standard sectioning command}
\documentclass[11pt,
  footinclude,headinclude,
  abstract=on
]{scrreprt}
%paper=b5

\input{preamble.tex}
\input{macros.tex}

\pagenumbering{roman}

\begin{document}

\input{metainfo.tex}

\cleardoublepage
\section*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

I would not be able to do this without plenty of luck and so many good people who helped me along the way.

I thank my family, especially my mom,
for all their love, support, and trust. Without you all, I would have never
embarked on this journey.

I dedicate this thesis to my late grandparents. Research took me far away
and I could not be with you, but I know you would be proud.

I thank my husband, Artem, for his endless love and care. Without you, I would not have finished this journey.

I thank my daughter, Sophia, for bringing me joy on tough days and giving me strength to go on.

I thank my advisor, Jan Vitek, not only for guiding me through this research journey, but also his kindness and support when it was most needed.

I thank my thesis committee, Amal Ahmed, Giuseppe Castagna, Arjun Guha, and Benjamin C. Pierce, for their helpful feedback, attention to detail, time, and flexibility.

I thank my first research advisor, Stanislav Mikhalkovich, for inviting me to programming languages research and supporting me.

I thank Ross Tate and Francesco Zappa Nardelli for their research mentorship.

I thank Vitaly Bragilevsky, Yana Demyanenko, and Jason Hemann for teaching with joy and making me feel that I belong.

I thank Matthias Felleisen for challenging and teaching me about teaching. 

I thank Jane Kokernak and Mitch Wand for teaching me about writing.

I thank my labmates and friends, especially Ellen Arteca, Benjamin Chung, Aviral Goel, Celeste Hollenbeck, Alexi Turcotte, and Ming-Ho Yee, for making this journey fun and less stressful.

I thank Alexandra Silva for mentorship and support during the most challenging part of this journey.

I thank all graduate worker parents out there who made me believe that having a baby during a PhD is possible.

I thank Chelsea, Sherisse, Terry, and Yiliana for taking care of Sophia while I was writing this thesis.


\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Contents}
\tableofcontents

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{\listfigurename}
\listoffigures

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{\listtablename}
\listoftables

%% ======================================================================
%% *********************************************************

% \section*{Proposal requirements}

% The thesis proposal should include the relevant background materials from
% literature and clearly specify the research problems to be attacked, the
% techniques to be used, and a schedule of milestones towards completion.
% Typically, the thesis proposal should not exceed 15 pages, excluding appendices
% and bibliography.

\cleardoublepage
\pagenumbering{arabic}

\input{chapters/chap-intro.tex}
%\input{chapters/chap-thesis.tex}
\input{chapters/chap-background.tex}
\input{chapters/chap-julia-sub.tex}
%\input{chapters/chap-4-prelim.tex}
%\input{chapters/chap-6-plan.tex}
\input{chapters/chap-dec-sub.tex}
\input{chapters/chap-evaluation.tex}
\input{chapters/chap-related.tex}
\input{chapters/chap-conclusion.tex}

\addcontentsline{toc}{chapter}{Bibliography}
\bibliography{bib/non-jb,bib/jb}

\appendix
\input{chapters/appendix.tex}

\end{document}
